---
title: "Assignment 5: Data Visualization"
author: "Shidi Dai"
output: pdf_document
geometry: margin=2.54cm
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct 14th @ 5:00pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse, 
lubridate, & cowplot packages. Upload the NTL-LTER processed data files for  
nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the
[`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()
#install.packages(tidyverse)
library(tidyverse)
#install.packages(lubridate)
library(lubridate)
#install.packages(cowpot)
library(cowplot)
NTL_LTER_Processed_Data <- 
  read.csv("../Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
Niwot_Ridge_Litter_Processed_Data <- 
  read.csv("../Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2 
class(NTL_LTER_Processed_Data$sampledate)
NTL_LTER_Processed_Data$sampledate <- 
  as.Date(NTL_LTER_Processed_Data$sampledate, format = "%Y-%m-%d")
class(Niwot_Ridge_Litter_Processed_Data$collectDate)
Niwot_Ridge_Litter_Processed_Data$collectDate <- 
  as.Date(Niwot_Ridge_Litter_Processed_Data$collectDate, format = "%Y-%m-%d")
```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and 
color it black. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r}
#4
plot04 <- 
  ggplot(NTL_LTER_Processed_Data, aes(x = po4, y = tp_ug, color = lakename)) +
  geom_point() +
  xlim(0, 30) +
  ylim(0, 75) +
  geom_smooth(method = lm, color = "black")
print(plot04)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;
see <https://r-lang.com/month-abb-in-r-with-example>

```{r}
#5
NTL_LTER_Processed_Data$month <- as.factor(NTL_LTER_Processed_Data$month)
factor(NTL_LTER_Processed_Data$month, levels = c(1:12))
#boxplot of temperature
Temperature_boxplot <-
  ggplot(NTL_LTER_Processed_Data, aes(x = month, y = temperature_C)) +
  geom_boxplot(aes(color = lakename))+
  scale_x_discrete(drop = FALSE)+
  theme(legend.position="none")
print(Temperature_boxplot)
#boxplot of TP
TP_boxplot <-
  ggplot(NTL_LTER_Processed_Data, aes(x = month, y = tp_ug)) +
  geom_boxplot(aes(color = lakename))+
  scale_x_discrete(drop = FALSE)+
  theme(legend.position="none")
print(TP_boxplot)
#boxplot of TN
TN_boxplot <-
  ggplot(NTL_LTER_Processed_Data, aes(x = month, y = tn_ug)) +
  geom_boxplot(aes(color = lakename))+
  scale_x_discrete(drop = FALSE)+
  theme(legend.position="bottom")
print(TN_boxplot)

library(cowplot)
plot_grid(Temperature_boxplot, TP_boxplot, TN_boxplot, nrow = 3, align = 'v', rel_heights = c(2, 2, 3))

```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer: Temperatures are higher in both lakes in the summer (July, August, September) and relatively lower before summer in May and June, and lower in November. The variable tp_ug is always higher in Peter Lake than in Paul lake from May till September. The variable tn_ug is also higher in Peter lake than in Paul lake from May to September. The difference in tn_ug between two lakes are bigger in July and August (summer). Tn_ug and tp_ug in the same lake are approximately the same throughout the year.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name 
of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r}
#6
Needles_plot <-
  ggplot(subset(Niwot_Ridge_Litter_Processed_Data, functionalGroup = "Needles"), 
                aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point()
print(Needles_plot)

#7
Needles_faceted <-
  ggplot(subset(Niwot_Ridge_Litter_Processed_Data, functionalGroup = "Needles"), 
                aes(x = collectDate, y = dryMass)) +
  geom_point()+
  facet_wrap("nlcdClass")
print(Needles_faceted)

Needles_faceted2 <-
  ggplot(subset(Niwot_Ridge_Litter_Processed_Data, functionalGroup = "Needles"), 
                aes(x = collectDate, y = dryMass)) +
  geom_point()+
  facet_grid("nlcdClass")
print(Needles_faceted2)
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:I think plot 7, my Needles_faceted plot is more effective. It separated the three nlcdclass, evergreenforest, grasslandherbaceous, and shrubscrub into three rows to show their distribution of drymass by collectdate. It is clearer than three colors messing together in the same graph (plot 6), which is hard to see the distribution of each nlcdclass. With plots in 7, we can easily compare the drymass and distributions between the three nlcdclasses.